Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-01-23 manu_draft.Rmd
through first attachment from JH

We describe and demonstre the proposed methods by analyzing water quality monitoring data from the southern portion of San Francisco Estuary, California, USA.  Approximately twice-monthly monitoring has been conducted for several decades at fixed locations (stations) on the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection which can generate artifacts affecting simple seasonal averages of the data.  We were interested in questions such as: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chlorophyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or baseline levels of productivity between blooms?  We provide examples illustrating how these questions can be addressed using methods with GAMs.  We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] after signal extraction with GAMs.  This approach is new to environmental trend-detection problems.  For this step, we provide methods using meta-anlaysis for isolating seasonal trends with reasonalbe certainty from from GAM results and evaluating these trends between years.

@@

The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly across years.  For example, if one year has an early and large magnitude spring bloom, the years before and after would have similar patterns described by the model, perhaps different but not substantially. For the raw SFE data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.  If the interaction spline has sufficient knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.  However, if that is the case, the conceptual motivation for model SYDI may not be justified.  In summary, model SDYI may best be suited for systems where within-year fluctuations actually vary smoothly across years, in which case it may out-perform the other models.  If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare (Table \@ref(tab:modstrtab)).

All of the models require a sufficiently large number of knots in one or more of the spline terms to accommodate the fluctuations in the data and use the GAM estimation methods in the manner for which they were designed.  In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.  In model SYDI, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.  Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.  In practice, allocation of knots between the different splines can be chosen by the analyst so long as it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.

@Murphy19 used a related set of spline formulations but used them differently than we do here. @Murphy19 have a "`gam0`" with only a `s(doy)` term, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.  Their `gam0` is *a priori* not of interest for our data and was not explored further.  Compared to our uses of SYD and SYDI, their uses differ in choice of knots and interpretation of results.  For SYD, they set a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years, whereas we use 12 times the number of years, roughly one per month.  Thus, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate signal from noise at a scale informed by the data.  In the SYDI model, @Murphy19 do not explicitly consider number of knots in the interaction spline.  In both cases, @Murphy19 in effect use choice of number of knots in different spline components to serve as ad hoc allocation of variation in the data to different components while not necessarily allowing any component to fully estimate the signal in the data. @Murphy19 acknowledge that incomplete modeling of fluctuations in the data may lead to inflated Type I error rates, which would apply to their later comparisons of changes across time, but they leave that problem for future work.  For the methods presented here, we seek to avoid inflated Type I error rates arising in this way.  Finally, @Murphy19 present large AIC differences between their spline formulations.  We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits (Table \@ref(tab:modstrtab)).  In our example, such large differences in AIC would only reflect inadequate choice of knots in one or more splines.

We present a simple visual comparison of chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially because of structural differences and differences in the penalties applied to the basis functions.  These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab). Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series, but the model predictions are very similar.

>>>>

We describe and demonstrate the proposed methods by analyzing water quality monitoring data from the southern portion of San Francisco Estuary, California, USA.  Approximately twice-monthly monitoring has been conducted for several decades at fixed locations (stations) on the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection which can generate artifacts affecting simple seasonal averages of the data.  We were interested in questions such as: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chlorophyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or baseline levels of productivity between blooms?  We provide examples illustrating how these questions can be addressed using methods with GAMs.  We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] after signal extraction with GAMs.  This approach is new to environmental trend-detection problems.  For this step, we provide methods using meta-anlaysis for isolating seasonal trends with reasonalbe certainty from from GAM results and evaluating these trends between years.

@@

The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year.  For example, if one year has an especially early and large magnitude spring bloom, the years before and after would be expected also to have earlier and larger than average spring blooms. For the raw SFE data, and chlorophyll-a dynamics in many estuaries, this is not expected to be the case. An early and large spring bloom could easily be followed the next year by a late and relatively small spring bloom.  If the interaction spline has many knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.  However, if that is the case, the conceptual motivation for model SYDI may not be supported by the data. Given that observed bloom dynamics for most systems do not vary smoothly across years, the potential benefits of using model SYDI could be outweighed by using simpler models.  Moreover, the spline and penalty structure differ from the other models, making it difficult to evaluate their relative performance (Table \@ref(tab:modstrtab)).

For models SYD and SYDI, the allocation of knots between the different splines can be chosen by the analyst so long as it is recognized that different choices will arbitrarily lead to related differences in allocation of variation among the splines. In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.  In model SYDI, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.  Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.

@Murphy19 used spline formulations related to those proposed here, but are insufficient for our needs of extracting and quantifying seasonal trends. @Murphy19 have a "`gam0`" with only a `s(doy)` term, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.  Their `gam0` is *a priori* not of interest for our data because of an assumption of constant seasonality and a linear annual trend.  Compared to our application of SYD and SYDI, their implementation of "`gam1`" and "`gam2`" allocated knots differently, leading to a different interpretation of results.  For "`gam1`" (our SYD), they set a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years, whereas we adopt a different approach using a number of knots equal to 12 times the number of years, or one per month.  Accordingly, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate intra-annual changes in season from noise at a finer temporal scale.  In the "`gam2`" model (our SYDI model), @Murphy19 do not explicitly consider number of knots in the interaction spline.  In both cases, they effectively use their choice of number of knots in different spline components as an ad hoc allocation of variation in the data to different components, while not allowing any component to fully estimate the signal in the data. @Murphy19 acknowledge that incomplete modeling of fluctuations in the data may inflate their Type I error rates for estimating temporal changes, but they leave that problem for future work.  We seek to avoid inflating Type I error rates in this way.  Finally, @Murphy19 present large AIC differences between their spline formulations.  We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits (Table \@ref(tab:modstrtab)), making differences in AIC unimportant.  In our example, large differences in AIC only reflect inadequate choice of knots in one or more splines, which should be avoided.

We visually compare chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).  Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.  Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).  However, when contrasting the models using only the continuous year smoother (`s(cont_year)`, as an example), the model fits differ substantially because of structural differences and differences in the penalties applied to the basis functions.  These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)). Accordingly, even though the models differ by which structural component describes variation in the chlorophyll time series, they provide similar predictions.
